Updated Constraints on Self-Interacting Dark Matter from Supernova 1987A

### 2017-05-10 draft.tex

An overwhelming preponderance of observational evidence indicates that a form of nonrelativistic, nonbaryonic, 
dark matter constitutes the majority of mass in the Universe and drives the formation of cosmic structure. 
The pace of the quest to identify the dark matter is accelerating on many fronts. Weakly-interacting massive 
particles (WIMPs) have received the most attention as dark matter candidates (see Ref.~\cite{jungman_etal96} for a review). 
Dark matter particles that interact with standard model particles only weakly, while interacting among themselves 
much more strongly have been studied as an alternative to WIMP scenarios in numerous contexts 
\cite{carlson_etal92,deLaix_etal95,atrio-barandela_davidson97,spergel_steinhardt00,hogan_dalcanton00,mohapatra_teplitz00,
dave_etal01,hisano_etal04,hisano_etal05,pospelov_etal08,arkani-hamed_etal08a,lattanzi_silk08,ackerman_etal09,feng_etal09,
kong_etal15} \arz{Cameron: Please review the literature and add some more recent references here. 
I know that Annika Peter has a few papers and Jonathan Feng probably has some more. 
Ensuring that you are up to date on these papers is important.} 
and constraints on self-interacting dark matter (SIDM) models have been explored by many authors \cite{yoshida_etal00,gnedin_ostriker01,miralda-escude02,randall_etal08,kamionkowski_profumo08,zentner09,robertson_zentner09,pieri_etal09,spolyar_etal09,finkbeiner_etal09,
slatyer_etal09,bramante_etal14,albuquerque_etal14,kaplinghat_etal14,chen_etal14,feng_etal16,catena_widmark16}. 
In this paper, we revisit and update astrophysical constraints on SIDM models from 
supernova cooling.

@@

We initiated our study because of a number of ambiguities appearing in the previous literature on this subject. 
In particular, we could not reproduce the constraints of Ref.~\cite{dent_etal12}. During the course of our 
study, we identified a number of errors in the analysis of Ref.~\cite{dent_etal12}. First, it is straightforward 
to demonstrate that the kinematical relationships given in Appendix A of Ref.~\cite{dent_etal12} are 
incorrect. Second, the squared matrix elements given in Eq.~(A3) and Eq.~(C18) of Ref.~\cite{dent_etal12} must 
be incorrect. 
%We have not been able to identify the errors which led to these incorrect expressions; however,  there must have been several oversights in the analysis of Ref.~\cite{dent_etal12}. 
For example, the squared 
matrix elements in Eq.~(A3) and Eq.~(C18) do not obey the correct symmetries under interchange of incoming 
and/or outgoing momenta. Furthermore, Ref.~\cite{dent_etal12} neglects the mass of the gauge boson, which 
is legitimate in the classic case of the $\sim$meV-mass axion, but not in the present context. Finally, Ref.~\cite{dent_etal12} 
employed an inconsistent model for the permitted energy loss rate from the supernova interior. Our work amounts 
primarily to repeating the calculation of Ref.~\cite{dent_etal12} in order to rectify these oversights. Our primary 
calculation treats nucleon scattering via one-pion exchange (OPE). We have repeated this analysis 
using an effective field theory for the nucleon interaction and find very similar results so we do not 
report on those in detail.

@@

The eight diagrams contribute the following eight terms to the pp amplitude, 
%
\begin{widetext}
\begin{eqnarray}
	M_1 &=& \frac{4 M_{\mathrm N}^2}{ m_\pi^2} \frac{f_{\mathrm{pp}}^2\, e\, \varepsilon}{k^2-m_\pi^2}  \frac{1}{m_{\mathrm{\Aprime}}^2 - 2 \qa \cdot p_1}\ 
	\bar{u}(p_4) \gamma_5 u(p_2)\  \bar{u}(p_3) \gamma_5 (\slashed{p}_1 - \qa +M_N)\slashed \epsilon  u(p_1), \\
	M_2 &=& -\frac{4 M_{\mathrm N}^2}{ m_\pi^2} \frac{f_{\mathrm{pp}}^2\, e\, \varepsilon}{l^2-m_\pi^2}  \frac{1}{m_{\mathrm{\Aprime}}^2 - 2 \qa \cdot p_1}\  
	\bar{u}(p_3) \gamma_5 u(p_2)\  \bar{u}(p_4) \gamma_5 (\slashed{p}_1 - \qa +M_N)\slashed{\epsilon} u(p_1),\\
	M_3 &=& \frac{4 M_{\mathrm N}^2}{ m_\pi^2} \frac{f_{\mathrm{pp}}^2\, e\, \varepsilon}{k^2-m_\pi^2}  \frac{1}{m_{\mathrm{\Aprime}}^2 - 2 \qa \cdot p_2}\ 
	\bar{u}(p_3) \gamma_5 u(p_1)\ \bar{u}(p_4) \gamma_5 (\slashed{p}_2 - \qa +M_N)\slashed \epsilon  u(p_2), \\
	M_4 &=& -\frac{4 M_{\mathrm N}^2}{ m_\pi^2} \frac{f_{\mathrm{pp}}^2\, e\, \varepsilon}{l^2-m_\pi^2}  \frac{1}{m_{\mathrm{\Aprime}}^2 - 2 \qa \cdot p_2}\ 
	\bar{u}(p_4) \gamma_5 u(p_1)\ \bar{u}(p_3) \gamma_5 (\slashed{p}_2 - \qa +M_N)\slashed \epsilon  u(p_2), \\
	M_5 &=& \frac{4 M_{\mathrm N}^2}{ m_\pi^2} \frac{f_{\mathrm{pp}}^2\, e\, \varepsilon}{k^2-m_\pi^2}  \frac{1}{m_{\mathrm{\Aprime}}^2 + 2 \qa \cdot p_3}\
	\bar{u}(p_4) \gamma_5 u(p_2)\ \bar{u}(p_3) \slashed{\epsilon} (\slashed{p}_3 + \qa +M_N)\gamma_5 u(p_1),\\
	M_6 &=& -\frac{4 M_{\mathrm N}^2}{ m_\pi^2} \frac{f_{\mathrm{pp}}^2\, e\, \varepsilon}{l^2-m_\pi^2}  \frac{1}{m_{\mathrm{\Aprime}}^2 + 2 \qa \cdot p_3}\ 
	\bar{u}(p_4) \gamma_5 u(p_1)\ \bar{u}(p_3) \slashed{\epsilon} (\slashed{p}_3 + \qa +M_N)\gamma_5 u(p_2),\\
	M_7 &=& \frac{4 M_{\mathrm N}^2}{ m_\pi^2} \frac{f_{\mathrm{pp}}^2\, e\, \varepsilon}{k^2-m_\pi^2}  \frac{1}{m_{\mathrm{\Aprime}}^2 + 2 \qa \cdot p_4}\ 
	\bar{u}(p_3) \gamma_5 u(p_1)\ \bar{u}(p_4) \slashed{\epsilon} (\slashed{p}_4 + \qa +M_N)\gamma_5 u(p_2),\\
	M_8 &=& -\frac{4 M_{\mathrm N}^2}{ m_\pi^2} \frac{f_{\mathrm{pp}}^2\, e\, \varepsilon}{l^2-m_\pi^2}  \frac{1}{m_{\mathrm{\Aprime}}^2 + 2 \qa \cdot p_4}\ 
	\bar{u}(p_3) \gamma_5 u(p_2)\ \bar{u}(p_4) \slashed{\epsilon} (\slashed{p}_4 + \qa +M_N)\gamma_5 u(p_1),
\end{eqnarray}
\end{widetext}
%
where the dark photon polarization is given by $\epsilon^{\nu}$. These expressions are identical to those 
given in Ref.~\cite{dent_etal12}; however, they do not simplify significantly if the correct kinematics are used. 
%Unfortunately, with the kinematical relations above, the squared amplitude does not yield a tidy expression for the spin-averaged squared matrix element. 
Our result contains over 200 terms, 
so we do not reproduce it here for reasons of convenience. However, we note that 
our result for $\vert \mathcal{M}_{\mathrm{pp}} \vert^2$ is symmetric 
under exchange of $k$ and $l$ as required.

@@

The quantity of interest is the rate of energy emission through dark gauge bosons. 
From the spin-summed, squared amplitudes described in the previous subsection, 
the energy emission rate is obtained by integrating over the phase space, 
and adding a factor of the energy of the emitted particle. To be specific, the 
energy emission rate per unit volume is 
%
\begin{widetext}
\beq
\label{eq:rate1}
Q_i = (2\pi)^4 \int\, E_{\mathrm{A}'} \, \sum_{s_1,s_2}\, \vert \mathcal{M}_i \vert^2 f(p_1) f(p_2)\delta^{(4)}(p_1+p_2-p_3-p_4-\qa)\, \dd \Pi,
\eeq
\end{widetext}
%
where 
%
\begin{equation}
d\Pi = \frac{\dd^3 \vec{q}_{\mathrm{A}'}}{(2\pi)^3 2E_{\mathrm{A}'}}\, \prod_{j=1}^{4}\, \frac{\dd^3 \vec{p}_j}{(2\pi^3) 2E_j}
\end{equation} 
is the Lorentz-invariant phase space interval, 
$ E_{\mathrm{A}'}$ is the energy of the emitted $\Aprime$ boson, 
$f(p)$ are the phase-space densities of the incoming nucleons, and 
the index $i$ on $Q_i$ refers to either the pp or pn processes. 
The nucleons in the core are comfortably non-degenerate and non-relativistic, 
so the Pauli blocking factor is omitted from Eq.~(\ref{eq:rate1}) and we take 
all nucleons to have a Maxwell-Boltzmann phase-space distribution distribution, 
\beq
f(p) =  \frac{n_{\mathrm{b}}}{2} \left( \frac{2 \pi}{M_{\mathrm{N}} T} \right)^{3/2} \exp \left(-\frac{p^2}{2 M_{\mathrm{N}} T} \right).
\eeq
%
We choose a baryon number density of $n_{\mathrm{b}} \approx 1.8 \times 10^{38}\, \mathrm{cm}^{-3}$ and a 
core supernova temperature of $T = 30\, \mathrm{MeV}$, both of which are typical choices and thought to 
be representative of supernova cores. \arz{Need to put a citation here for the properties of supernova cores. 
I can probably find something in the literature, but please put something here if you have a citation.} \cbm{On second thought, is Raffelt not good enough for both of these? That's where everybody started}

@@

To obtain the dark gauge boson luminosity from $Q_{\mathrm{pp}}$ and $Q_{\mathrm{pn}}$, 
we assumed that $\Aprime$ production takes place in a stellar core of radius $\sim 1$~km, 
so that the total luminosity of $\Aprime$ is 
%
\begin{equation}
L_{\mathrm{A}'} = V (Q_{\mathrm{pp}} + Q_{\mathrm{pn}}),
\end{equation}
%
where $V$ is the volume of the sphere, and the luminosity per unit mass within the core is 
%
\begin{equation}
\left( \frac{L_{\mathrm{A'}}}{M} \right) = \frac{Q_{\mathrm{pp}} + Q_{\mathrm{pn}}}{\rho}, 
\end{equation}
%
where $\rho = 3 \times 10^{14}\, \mathrm{g/cm}^3$ is the mass density of the core\footnote{
We use the symbol $\left( L_{\mathrm{A'}}/M \right)$ for the luminosity per unit mass rather than 
the more conventional symbol $\epsilon_{\mathrm{A'}}$ in order to avoid confusion with the 
mixing parameter $\varepsilon$.}. Following previous studies, the energy loss rate into novel 
particles cannot exceed 
%
\begin{equation}
\left( \frac{L_{\mathrm{max}}}{M} \right) \approx 10^{19} \, \frac{\mathrm{erg}}{\mathrm{g}\cdot \mathrm{s}}
\end{equation}
%
without significantly reducing the duration of the neutrino burst observed at Earth for SN1987A 
\cite{raffelt96_book}. Writing $(L_{\mathrm{A'}}/M) = \varepsilon^2 I_{\mathrm{A'}}(m_{\mathrm{A'}},T)$, 
we delineate the constraint on the dark photon mixing parameter by 
%
\begin{equation}
\label{eq:bound1}
\varepsilon \lesssim \sqrt{\frac{(L_{\mathrm{max}}/M)}{I_{\mathrm{A}'}(m_{\mathrm{A}'}, T)}}.
\end{equation}
%
At this point, we note that in Ref.~\cite{dent_etal12} took a value of 
$(L_{\mathrm{max}}/M)$ approximately three orders of magnitude larger than 
this accepted value. Remarkably, this nearly canceled the errors in their evaluation of 
$Q_{\mathrm{pp}}$ and $Q_{\mathrm{np}}$ so that they quote constraints that are 
within an order of magnitude of the correct result.

@@

\arz{This section needs some work as it is a bit hard to follow.} 
\cbm{I think I've spent too long dealing with the calculation to fix this by myself. I can tell it's a bit confusing as a whole, but when I focus on any one part it of course all pretty much makes sense
	to me}

The second effect that produces an upper bound on the excluded region 
comes from considering trapping of dark bosons within the supernova. 
With a large enough coupling, the dark photons will thermalize and then 
will be emitted from an approximately spherical ``dark photosphere" at the radial position 
where the $\Aprime$ mean free path becomes larger than the typical size of the supernova core. 
In this case the luminosity is given simply by Stefan's law, 
\beq
L_{\mathrm{A'}}  = 4\pi \rdp^2 T_{\mathrm{A'}}^4 \sigma,
\eeq
where $\rdp$ is now the radius of the emitting shell and $T_{\mathrm{A'}}$ its temperature. 
We estimate the radius of this dark photosphere as $\rdp=10$ km, 
because the density of the supernova drops drastically around that point. 
\arz{We need a citation for this statement, but I do not know one off of the top of my head.}
The bound on the luminosity can then be recast as a bound on $T_{\mathrm{A}'}$, 
%
\beq
T_{\mathrm{A}'} \lesssim 9.6\rm\ MeV.
\eeq
%
That bound can then be translated into the desired bound on the coupling as a function of mass by adopting a 
simple model for the supernova atmosphere, assuming that the particles are emitted from the dark photosphere 
at a point where the optical depth to scattering is $\tau = 2/3 $, and finding the temperature that corresponds 
to that optical depth.

This is a somewhat involved calculation. First, one needs a model for 
the density and temperature in the supernova. Following \cite{turner88}, 
we assume a simple power-law model for the supernova core, with 
\bea
\rho &=& \rho_{\mathrm{p}} \left( \frac{R}{r}\right )^n,\\
T &=& T_{\mathrm{R}} \left[ \frac{\rho(r)}{\rho_{\mathrm{p}}} \right]^{1/3},
\eea
with $  \rho_{\mathrm{p}}  = 3\times10^{14}\ {\rm g/cm}^3$, $T_{\mathrm{R}} = 30$~MeV, $R = 10$ km,
and $n = 5$.
The optical depth is given by 
\beq
\tau = \int_{r_x}^{\infty}\, \kappa \rho\,  \dd r,
\eeq
where $ \kappa$ is the opacity, which we take to be the Rosseland mean 
%
\beq 
\frac{1}{\kappa \rho} = \frac{15}{4 \pi^4 T^5}\, \int_{M_{\mathrm{A'}}}^{\infty}\, 
\frac{E_{\mathrm{A'}}^2\, e^{E_{\mathrm{A'}}/T} 
\sqrt{E_{\mathrm{A'}}^2 - m_{\mathrm{A'}}^2}}
{(e^{E_{\mathrm{A'}}/T}-1)^2}\, l_{\mathrm{A'}}\, \dd E_{\mathrm{A'}},
\eeq
where $l_{\mathrm{A'}}$ is the mean free path.

The inverse mean free path can readily be obtained by modifying $ Q_i $, 
the expression for the energy loss rate, as follows: removing the factor of $ E_\mathrm{A'} $ 
and the phase space integral over $ \qa $, and adding a factor of $ e^{E_\mathrm{A'}/T} $ for detailed balance. 
\arz{I'm not sure that I follow this somewhat cavalier statement about detailed balance. Can someone 
please explain this to me when we meet. In my earlier rounds of thinking about this, I took that factor 
at face value as given, assuming I'd eventually be able to figure out where it comes from.} 
This gives the inverse mean free path as a function of mass and coupling. 
Again, we perform the required integration numerically using the {\tt CUBA }package. 
This then allows the calculation of $\kappa_{\mathrm{pp}}$ and $\kappa_{\mathrm{pn}}$, 
which are the opacities due to inverse bremsstrahlung for proton-proton and proton-neutron 
processes respectively. These dominate the opacity of the star to $A'$ propagation. 
The inverse opacities for the $pn$ and $pp$ processes add, 
giving the total opacity $ \kappa^{-1} = \kappa_{\mathrm{pp}}^{-1} + \kappa_{\mathrm{pn}}^{-1} $

Having obtained an expression for $ \kappa $, we can now find the optical depth as follows. 
Define a typical optical depth as $\tau_\mathrm{R} = \kappa_\mathrm{R} \rho_\mathrm{R} R $. We then have 
\beq 
\kappa \rho R = \tau_\mathrm{R} \left( \frac{\rho}{\rho_\mathrm{R}} \right)^2 \left( \frac{T_\mathrm{R}}{T} \right)^{3/2}.
\eeq
%
This can be combined with the previous expressions for the density and temperature as a function of position 
and plugged in to the integral expression for the optical depth to obtain 
\bea 
\tau &=& \int_{r_x}^{\infty}\, \tau_\mathrm{R} \left( \frac{R}{r} \right)^{3n/2} \dd r \\
 &=& \frac{\tau_\mathrm{r}}{\frac{3n}{2}-1} \left( \frac{T_\mathrm{A'}}{T_\mathrm{R}} \right)^{(9/2-3/n)}. 
 \eea
 The bound on the coupling is obtained by requiring  $ \tau(\varepsilon,m_\mathrm{A'}) \le 2/3 $

 \arz{It seems as though you can check the consistency of your assumptions can 
 be checked within this calculation. For example, you can actually use this calculation 
 to estimate the radius at which $\tau=2/3$ and see if it is actually close to 
 $10$~km. Have you done this?} %On second thought, put that separately%
\akl{Cameron, you did this correct?  Can you describe?}

@@

\cbm{Should the dependence on specific temperature chosen - viz. that it's at most a roughly comparable factor - be mentioned here?}

>>>>

An overwhelming preponderance of observational evidence indicates that a form of nonrelativistic, nonbaryonic, 
dark matter constitutes the majority of mass in the Universe and drives the formation of cosmic structure. 
The pace of the quest to identify the dark matter is accelerating on many fronts. Weakly-interacting massive 
particles (WIMPs) have received the most attention as dark matter candidates (see Ref.~\cite{jungman_etal96} for a review). 
Dark matter particles that interact with standard model particles only weakly, while interacting among themselves 
much more strongly have been studied as an alternative to WIMP scenarios in numerous contexts 
\cite{carlson_etal92,deLaix_etal95,atrio-barandela_davidson97,spergel_steinhardt00,hogan_dalcanton00,mohapatra_teplitz00,
dave_etal01,hisano_etal04,hisano_etal05,pospelov_etal08,arkani-hamed_etal08a,lattanzi_silk08,ackerman_etal09,feng_etal09,
kong_etal15} \arz{Cameron: Please review the literature and add some more recent references here. 
I know that Annika Peter has a few papers and Jonathan Feng probably has some more. 
Ensuring that you are up to date on these papers is important.} \arz{As far as I can see, no new 
references were added here or in any other relevant spots in the paper. Please do a literature search. At the very least, 
I know of papers by Annika Peter and Jonathan Feng that may be relevant here.}
and constraints on self-interacting dark matter (SIDM) models have been explored by many authors \cite{yoshida_etal00,gnedin_ostriker01,miralda-escude02,randall_etal08,kamionkowski_profumo08,zentner09,robertson_zentner09,pieri_etal09,spolyar_etal09,finkbeiner_etal09,slatyer_etal09,bramante_etal14,albuquerque_etal14,kaplinghat_etal14,chen_etal14,feng_etal16,catena_widmark16}. In this paper, we revisit and update astrophysical constraints on SIDM models from supernova cooling.

@@

We initiated our study because of a number of ambiguities appearing in the previous literature on this subject. 
In particular, we could not reproduce the constraints of Ref.~\cite{dent_etal12}. During the course of our 
study, we identified a number of errors in the analysis of Ref.~\cite{dent_etal12}. First, it is straightforward 
to demonstrate that the kinematical relationships given in Appendix A of Ref.~\cite{dent_etal12} are 
incorrect. Second, the squared matrix elements given in Eq.~(A3) and Eq.~(C18) of Ref.~\cite{dent_etal12} must 
be incorrect. These matrix elements do not obey the correct symmetries under interchange of incoming 
and/or outgoing momenta. Furthermore, Ref.~\cite{dent_etal12} neglects the mass of the gauge boson, which 
is legitimate in the classic case of the $\sim$meV-mass axion, but not in the present context. Finally, Ref.~\cite{dent_etal12} 
employed an inconsistent model for the permitted energy loss rate from the supernova interior. Our work amounts 
primarily to repeating the calculation of Ref.~\cite{dent_etal12} in order to rectify these oversights. Our primary 
calculation treats nucleon scattering via one-pion exchange (OPE). We have repeated this analysis 
using an effective field theory for the nucleon interaction and find very similar results so we do not 
report on those in detail.

@@

The eight diagrams contribute the following eight terms to the pp amplitude, 
%
\begin{widetext}
\begin{eqnarray}
	M_1 &=& \frac{4 M_{\mathrm N}^2}{ m_\pi^2} \frac{f_{\mathrm{pp}}^2\, e\, \varepsilon}{k^2-m_\pi^2}  \frac{1}{m_{\mathrm{\Aprime}}^2 - 2 \qa \cdot p_1}\ 
	\bar{u}(p_4) \gamma_5 u(p_2)\  \bar{u}(p_3) \gamma_5 (\slashed{p}_1 - \qa +M_N)\slashed \epsilon  u(p_1), \\
	M_2 &=& -\frac{4 M_{\mathrm N}^2}{ m_\pi^2} \frac{f_{\mathrm{pp}}^2\, e\, \varepsilon}{l^2-m_\pi^2}  \frac{1}{m_{\mathrm{\Aprime}}^2 - 2 \qa \cdot p_1}\  
	\bar{u}(p_3) \gamma_5 u(p_2)\  \bar{u}(p_4) \gamma_5 (\slashed{p}_1 - \qa +M_N)\slashed{\epsilon} u(p_1),\\
	M_3 &=& \frac{4 M_{\mathrm N}^2}{ m_\pi^2} \frac{f_{\mathrm{pp}}^2\, e\, \varepsilon}{k^2-m_\pi^2}  \frac{1}{m_{\mathrm{\Aprime}}^2 - 2 \qa \cdot p_2}\ 
	\bar{u}(p_3) \gamma_5 u(p_1)\ \bar{u}(p_4) \gamma_5 (\slashed{p}_2 - \qa +M_N)\slashed \epsilon  u(p_2), \\
	M_4 &=& -\frac{4 M_{\mathrm N}^2}{ m_\pi^2} \frac{f_{\mathrm{pp}}^2\, e\, \varepsilon}{l^2-m_\pi^2}  \frac{1}{m_{\mathrm{\Aprime}}^2 - 2 \qa \cdot p_2}\ 
	\bar{u}(p_4) \gamma_5 u(p_1)\ \bar{u}(p_3) \gamma_5 (\slashed{p}_2 - \qa +M_N)\slashed \epsilon  u(p_2), \\
	M_5 &=& \frac{4 M_{\mathrm N}^2}{ m_\pi^2} \frac{f_{\mathrm{pp}}^2\, e\, \varepsilon}{k^2-m_\pi^2}  \frac{1}{m_{\mathrm{\Aprime}}^2 + 2 \qa \cdot p_3}\
	\bar{u}(p_4) \gamma_5 u(p_2)\ \bar{u}(p_3) \slashed{\epsilon} (\slashed{p}_3 + \qa +M_N)\gamma_5 u(p_1),\\
	M_6 &=& -\frac{4 M_{\mathrm N}^2}{ m_\pi^2} \frac{f_{\mathrm{pp}}^2\, e\, \varepsilon}{l^2-m_\pi^2}  \frac{1}{m_{\mathrm{\Aprime}}^2 + 2 \qa \cdot p_3}\ 
	\bar{u}(p_4) \gamma_5 u(p_1)\ \bar{u}(p_3) \slashed{\epsilon} (\slashed{p}_3 + \qa +M_N)\gamma_5 u(p_2),\\
	M_7 &=& \frac{4 M_{\mathrm N}^2}{ m_\pi^2} \frac{f_{\mathrm{pp}}^2\, e\, \varepsilon}{k^2-m_\pi^2}  \frac{1}{m_{\mathrm{\Aprime}}^2 + 2 \qa \cdot p_4}\ 
	\bar{u}(p_3) \gamma_5 u(p_1)\ \bar{u}(p_4) \slashed{\epsilon} (\slashed{p}_4 + \qa +M_N)\gamma_5 u(p_2),\\
	M_8 &=& -\frac{4 M_{\mathrm N}^2}{ m_\pi^2} \frac{f_{\mathrm{pp}}^2\, e\, \varepsilon}{l^2-m_\pi^2}  \frac{1}{m_{\mathrm{\Aprime}}^2 + 2 \qa \cdot p_4}\ 
	\bar{u}(p_3) \gamma_5 u(p_2)\ \bar{u}(p_4) \slashed{\epsilon} (\slashed{p}_4 + \qa +M_N)\gamma_5 u(p_1),
\end{eqnarray}
\end{widetext}
%
where the dark photon polarization is given by $\epsilon^{\nu}$. These expressions are identical to those 
given in Ref.~\cite{dent_etal12}; however, they do not simplify significantly if the correct kinematics are used. 
Our result contains over 200 terms, so we do not reproduce it here for reasons of convenience. However, we note that our result for $\vert \mathcal{M}_{\mathrm{pp}} \vert^2$ is symmetric under exchange of $k$ and $l$ as required.

@@

The quantity of interest is the rate of energy emission through dark gauge bosons. 
From the spin-summed, squared amplitudes described in the previous subsection, 
the energy emission rate is obtained by integrating over the phase space, 
and adding a factor of the energy of the emitted particle. To be specific, the 
energy emission rate per unit volume is 
%
\begin{widetext}
\beq
\label{eq:rate1}
Q_i = (2\pi)^4 \int\, E_{\mathrm{A}'} \, \sum_{s_1,s_2}\, \vert \mathcal{M}_i \vert^2 f(p_1) f(p_2)\delta^{(4)}(p_1+p_2-p_3-p_4-\qa)\, \dd \Pi,
\eeq
\end{widetext}
%
where 
%
\begin{equation}
d\Pi = \frac{\dd^3 \vec{q}_{\mathrm{A}'}}{(2\pi)^3 2E_{\mathrm{A}'}}\, \prod_{j=1}^{4}\, \frac{\dd^3 \vec{p}_j}{(2\pi^3) 2E_j}
\end{equation} 
is the Lorentz-invariant phase space interval, 
$ E_{\mathrm{A}'}$ is the energy of the emitted $\Aprime$ boson, 
$f(p)$ are the phase-space densities of the incoming nucleons, and 
the index $i$ on $Q_i$ refers to either the pp or pn processes. 
The nucleons in the core are comfortably non-degenerate and non-relativistic, 
so the Pauli blocking factor is omitted from Eq.~(\ref{eq:rate1}) and we take 
all nucleons to have a Maxwell-Boltzmann phase-space distribution distribution, 
\beq
f(p) =  \frac{n_{\mathrm{b}}}{2} \left( \frac{2 \pi}{M_{\mathrm{N}} T} \right)^{3/2} \exp \left(-\frac{p^2}{2 M_{\mathrm{N}} T} \right).
\eeq
%
We choose a baryon number density of $n_{\mathrm{b}} \approx 1.8 \times 10^{38}\, \mathrm{cm}^{-3}$ and a 
core supernova temperature of $T = 30\, \mathrm{MeV}$, both of which are typical choices and thought to 
be representative of supernova cores.

@@

To obtain the dark gauge boson luminosity from $Q_{\mathrm{pp}}$ and $Q_{\mathrm{pn}}$, 
we assumed that $\Aprime$ production takes place in a stellar core of radius $\sim 1$~km, 
so that the total luminosity of $\Aprime$ is 
%
\begin{equation}
L_{\mathrm{A}'} = V (Q_{\mathrm{pp}} + Q_{\mathrm{pn}}),
\end{equation}
%
where $V$ is the volume of the sphere, and the luminosity per unit mass within the core is 
%
\begin{equation}
\left( \frac{L_{\mathrm{A'}}}{M} \right) = \frac{Q_{\mathrm{pp}} + Q_{\mathrm{pn}}}{\rho}, 
\end{equation}
%
where $\rho = 3 \times 10^{14}\, \mathrm{g/cm}^3$ is the mass density of the core\footnote{
We use the symbol $\left( L_{\mathrm{A'}}/M \right)$ for the luminosity per unit mass rather than 
the more conventional symbol $\epsilon_{\mathrm{A'}}$ in order to avoid confusion with the 
mixing parameter $\varepsilon$.}. Following previous studies, the energy loss rate into novel 
particles cannot exceed 
%
\begin{equation}
\left( \frac{L_{\mathrm{max}}}{M} \right) \approx 10^{19} \, \frac{\mathrm{erg}}{\mathrm{g}\cdot \mathrm{s}}
\end{equation}
%
without significantly reducing the duration of the neutrino burst observed at Earth for SN1987A 
\cite{raffelt96_book}. Writing $(L_{\mathrm{A'}}/M) = \varepsilon^2 I_{\mathrm{A'}}(m_{\mathrm{A'}},T)$, 
we delineate the constraint on the dark photon mixing parameter by 
%
\begin{equation}
\label{eq:bound1}
\varepsilon \lesssim \sqrt{\frac{(L_{\mathrm{max}}/M)}{I_{\mathrm{A}'}(m_{\mathrm{A}'}, T)}}.
\end{equation}
%
At this point, we note that Dent et al. in Ref.~\cite{dent_etal12} took a value of 
$(L_{\mathrm{max}}/M)$ approximately three orders of magnitude larger than 
this accepted value. Remarkably, this nearly canceled the errors in their evaluation of 
$Q_{\mathrm{pp}}$ and $Q_{\mathrm{np}}$ so that they quote constraints that are 
within an order of magnitude of the correct result.

@@

The second effect that produces an upper bound on the excluded region 
comes from considering trapping of dark bosons within the supernova. 
With a large enough coupling, the dark photons will thermalize and then 
will be emitted from an approximately spherical ``dark photosphere" at the radial position 
where the $\Aprime$ mean free path becomes larger than the typical size of the supernova core. 
In this case the luminosity is given simply by Stefan's law, 
\beq
L_{\mathrm{A'}}  = 4\pi \rdp^2 T_{\mathrm{A'}}^4 \sigma,
\eeq
where $\rdp$ is now the radius of the emitting shell and $T_{\mathrm{A'}}$ its temperature. 
We estimate the radius of this dark photosphere as $\rdp=10$ km, 
because the density of the supernova drops drastically around that point. We will 
confirm shortly that this estimate is consistent within the context of the simple model that we 
adopt for the interior structure of the supernova atmospherer. The bound on the luminosity can then 
be recast as a bound on $T_{\mathrm{A}'}$, 
%
\beq
T_{\mathrm{A}'} \lesssim 9.6\rm\ MeV.
\eeq
%
That bound can then be translated into the desired bound on the coupling as a function of mass by adopting a 
simple model for the supernova atmosphere, assuming that the particles are emitted from the dark photosphere 
at a point where the optical depth to scattering is $\tau = 2/3 $, and finding the temperature that corresponds 
to that optical depth.

This is a somewhat involved calculation. First, one needs a model for the density and temperature in the supernova. Following the simple, early model of Ref.~\cite{turner88}, we assume a simple power-law model for the supernova core, with 
\bea
\rho &=& \rho_{\mathrm{p}} \left( \frac{R}{r}\right )^n,\\
T &=& T_{\mathrm{R}} \left[ \frac{\rho(r)}{\rho_{\mathrm{p}}} \right]^{1/3},
\eea
with $  \rho_{\mathrm{p}}  = 3\times10^{14}\ {\rm g/cm}^3$, $T_{\mathrm{R}} = 30$~MeV, $R = 10$ km,
and $n = 5$.
The optical depth is given by 
\beq
\tau = \int_{r_x}^{\infty}\, \kappa \rho\,  \dd r,
\eeq
where $ \kappa$ is the opacity, which we take to be the Rosseland mean 
%
\beq 
\frac{1}{\kappa \rho} = \frac{15}{4 \pi^4 T^5}\, \int_{M_{\mathrm{A'}}}^{\infty}\, 
\frac{E_{\mathrm{A'}}^2\, e^{E_{\mathrm{A'}}/T} 
\sqrt{E_{\mathrm{A'}}^2 - m_{\mathrm{A'}}^2}}
{(e^{E_{\mathrm{A'}}/T}-1)^2}\, l_{\mathrm{A'}}\, \dd E_{\mathrm{A'}},
\eeq
where $l_{\mathrm{A'}}$ is the mean free path.

The inverse mean free path can be obtained by modifying $ Q_i $, 
the expression for the energy loss rate, as follows: removing the factor of $ E_\mathrm{A'} $ 
and the phase space integral over $ \qa $, and adding a factor of $ e^{E_\mathrm{A'}/T} $ for detailed balance. 
\arz{I'm not sure that I follow this somewhat cavalier statement about detailed balance. Can someone 
please explain this to me when we meet. In my earlier rounds of thinking about this, I took that factor 
at face value as given, assuming I'd eventually be able to figure out where it comes from.} 
This gives the inverse mean free path as a function of mass and coupling. 
Again, we perform the required integration numerically using the {\tt CUBA }package. 
This then allows the calculation of $\kappa_{\mathrm{pp}}$ and $\kappa_{\mathrm{pn}}$, 
which are the opacities due to inverse bremsstrahlung for proton-proton and proton-neutron 
processes respectively. These dominate the opacity of the star to $A'$ propagation. 
The inverse opacities for the $pn$ and $pp$ processes add, 
giving the total opacity $ \kappa^{-1} = \kappa_{\mathrm{pp}}^{-1} + \kappa_{\mathrm{pn}}^{-1} $

Having obtained an expression for $ \kappa $, we can now find the optical depth as follows. 
Define a typical optical depth as $\tau_\mathrm{R} = \kappa_\mathrm{R} \rho_\mathrm{R} R $. We then have 
\beq 
\kappa \rho R = \tau_\mathrm{R} \left( \frac{\rho}{\rho_\mathrm{R}} \right)^2 \left( \frac{T_\mathrm{R}}{T} \right)^{3/2}.
\eeq
%
This can be combined with the previous expressions for the density and temperature as a function of position 
and plugged in to the integral expression for the optical depth to obtain 
\bea 
\tau &=& \int_{r_x}^{\infty}\, \tau_\mathrm{R} \left( \frac{R}{r} \right)^{3n/2} \dd r \\
 &=& \frac{\tau_\mathrm{r}}{\frac{3n}{2}-1} \left( \frac{T_\mathrm{A'}}{T_\mathrm{R}} \right)^{(9/2-3/n)}. 
 \eea
 The bound on the coupling is obtained by requiring  $ \tau(\varepsilon,m_\mathrm{A'}) \le 2/3 $. Note that strictly speaking we should have determined $ r_{\mathrm{dp}}$ rather than assuming a value. However, it is possible to verify the self-consistency of our assumption using this model for the supernova atmosphere. In particular, this model implies that an optical depth of $\tau=2/3$ is reached at a radial position of $ r_{\mathrm{dp}}=XXXX$,  validating the assumption made at the outset. \arz{Put in the missing number here.}

@@
